Measuring contrast processing in the visual system using SSVEP

### 2024-04-06 review.qmd
Started clinical section

---
title: "Measuring contrast processing in the visual system using SSVEP"
format: pdf
bibliography: WadeBaker_Review.bib
csl: visual-neuroscience.csl
execute:
  echo: false
---

## Wade, A.R. & Baker, D.H.: Review for Vis Neurosci

## ABSTRACT:

@@

## INTRODUCTION

@@

The SSVEP is a continuous electrical response evoked in the brain by visual stimuli flickering at a constant frequency [@Regan1966]. For contrast-defined stimuli, such as sine-wave gratings, it is strongest at the occipital pole, adjacent to the early visual areas that generate the signal, although careful analysis of individual VEPs reveals multiple generators throughout visual cortex [@Di-Russo2007] @Hillebrand. The flickering stimulus entrains neural population responses at multiples of the stimulus frequency, so continuous EEG data are typically analysed by taking the Fourier transform, and estimating the amplitude at these frequencies. Two common variants involve sinusoidal on-off flicker, where the stimulus alternates between a blank background and the peak contrast, and sinusoidal counterphase flicker, where the stimulus alternates in phase (i.e the black regions become white and the white regions become black). On-off flicker can drive independent populations of on- and off-cells in the retina once per cycle and can therefore produce a response at the fundamental flicker frequency, known as 1F, and its integer harmonics: 2F, 3F, 4F and so on. Counterphase flicker contains two transients per cycle and therefore does not produce a response at 1F, only at its even harmonics: 2F, 4F, 6F and so on. Because square-waves are spectrally broad-band, square wave flicker tends to produce more complex spectral harmonics than sine-wave flicker.

@@

### Why measure responses to contrast?

@@

This possibility was recognised in the early days of the technique [@Regan1988]. Although characterising the complete set of computations along entire processing pathway is problematic, careful parametric variation of the input stimuli does allow us to fit models of early visual gain control using SSVEP data. ... MORE HERE - Although this idea was suggested as early as 1988 (probably earlier) not many people seem to have done it much. @Candy2001; @Busse2009; @Tsai2012; @Baker2017 Focusing only on SSVEP: [@NorciaCandy2001, @Regan1990]

### Measuring the development of contrast processing

@@

##

>>>>

---
title: "Measuring contrast processing in the visual system using SSVEP"
author:
  - name: Alex R. Wade
  - name: Daniel H. Baker
format: 
  pdf:
    keep-tex: true
bibliography: WadeBaker_Review.bib
csl: visual-neuroscience.csl
execute:
  echo: false
---
<p style="text-align: center;">Department of Psychology and York Biomedical Research Institute

University of York, UK</p>

## Review for Vis Neurosci

# Abstract:

@@

# Introduction

@@

The SSVEP is a continuous electrical response evoked in the brain by visual stimuli flickering at a constant frequency [@Regan1966]. For contrast-defined stimuli, such as sine-wave gratings, it is strongest at the occipital pole, adjacent to the early visual areas that generate the signal, although careful analysis of individual VEPs reveals multiple generators throughout visual cortex [@Di-Russo2007]. The flickering stimulus entrains neural population responses at multiples of the stimulus frequency, so continuous EEG data are typically analysed by taking the Fourier transform, and estimating the amplitude at these frequencies. Two common variants involve sinusoidal on-off flicker, where the stimulus alternates between a blank background and the peak contrast, and sinusoidal counterphase flicker, where the stimulus alternates in phase (i.e the black regions become white and the white regions become black). On-off flicker can drive independent populations of on- and off-cells in the retina once per cycle and can therefore produce a response at the fundamental flicker frequency, known as 1F, and its integer harmonics: 2F, 3F, 4F and so on. Counterphase flicker contains two transients per cycle and therefore does not produce a response at 1F, only at its even harmonics: 2F, 4F, 6F and so on. Because square-waves are spectrally broad-band, square wave flicker tends to produce more complex spectral harmonics than sine-wave flicker.

@@

# Why measure responses to contrast?

@@

This possibility was recognised in the early days of the technique [@Regan1988]. Although characterising the complete set of computations along entire processing pathway is problematic, careful parametric variation of the input stimuli does allow us to fit models of early visual gain control using SSVEP data. ... MORE HERE - Although this idea was suggested as early as 1988 (probably earlier) not many people seem to have done it much. @Candy2001; @Busse2009; @Tsai2012; @Baker2017

# Measuring the development of contrast processing

@@

# Contrast processing in clinical conditions

The SSVEP technique has also been used to study clinical conditions, such as diseases and developmental disorders. This can often be informative regarding the underlying mechanism that characterises the condition. Here we focus on four conditions, but there is potential to apply the method more broadly, either as a diagnostic technique, or to monitor disease severity and progression, or the efficacy of treatments.

Epilepsy is a neurological condition in which patients experience seizures - episodes of uncontrolled neural activity that can cause unconsciousness, involuntary movements and convulsions, and atypical sensory experiences. @Porciatti2000 showed that individuals with photosensitive epilepsy generate larger steady-state signals in response to flickering visual stimuli, and their contrast response functions saturate less than those of healthy controls. This is consistent with the idea that epilepsy involves a cortical hyperexciteability that makes seizures more likely. It is also the case for individuals with idiopathic generalised epilepsy [@Tsai2011], a subtype of epilepsy that has a less obvious link to vision. The differences apply across the whole contrast-response function, and so resemble a response gain effect, which might be due to reduced inhibition from neighbouring neurons. Differences in SSVEP amplitudes have also been reported in individuals with migraine [@Shibata2008], a condition also associated with cortical hyperexciteability.

Amblyopia

Autism

Parkinson's

# Brain-computer interfaces

# Conclusions
